NumPy---Array Computation for Python

### 2020-01-10 body.tex
Add missing link
Another pass through array structure

\begin{figure*}
  \centering
  \includegraphics[width=\textwidth]{static/sketches/array-concepts}
  \caption{\fixme{PLACEHOLDER: This figure needs some text to describe
  it}
   }
  \label{fig:arrayconcepts}
\end{figure*}

@@

The NumPy array consists of a pointer to memory, along with meta-data used to
interpret the data stored there.
For example, while computer memory stores elements linearly and has no notion
of higher-dimensions, we can interpret it as $n$-dimensional arrays by also
providing its shape, strides (the number of bytes to advance in memory to skip
along rows, columns, etc.), and data-type.

Data types (dtypes for short) tell us more about the elements stored in a NumPy
array.
Arrays can hold, e.g., numerical data (of lower and higher precision), strings,
datetimes, or Python objects, each identified by its own dtype.

Once data is viewed as an $n$-dimensional array, more abstract operations such
as reductions (e.g., summation across columns), products, and so forth, are
defined (Figure~\ref{fig:reductions}).
Users predominantly interact with NumPy arrays using these operators and
associated utility functions through an easily readable high-level API
(Application Programming Interface).

Sometimes, it is desirable to combine arrays of different dimensions.
One may, for example, want to add a scalar to a vector, or scale each column of
an array by a specific scalar value.
This is done through {\em broadcasting}, where one of the arrays is virtually
duplicated (i.e., without copying any data in memory), so that the shapes of
the operands match.
Broadcasting is illustrated in Figure~\ref{fig:broadcasting}.

@@

NumPy include numerous functions that perform element-wise and reduction
operations on arrays, including arithmetic, statistics, and trigonometry.
They aim to loop over array elements near-optimally (taking into consideration,
e.g., strides, in order to best utilize the computer's fast cache memory), and
implement broadcasting.
Vectorized operations that would take many tens of lines to express in
languages such as C can often be implemented as a single, clear Python
expression.

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{static/sketches/reductions}
  \caption{\fixme{Reducing an array along one of several axes.}
   }
  \label{fig:reductions}
\end{figure}

NumPy also provides support for accelerated linear algebra, and supports
several established backends such as OpenBLAS and Intel MKL.
Suppose the matrix \code{A} is a 2-d array and the vector \code{x} is a 1-d array.
Then \code{A @ x} computes the matrix-vector product $Ax$ and \code{x @ A} computes
the vector-matrix product $x^\top A$.
This matrix multiplication operation extends to higher-dimensions by treating
the array as a stack of matrices.

In addition, NumPy provides a large collection of utility functions for
creating, reshaping, concatenating, and padding arrays; for searching, sorting
and counting data in arrays; and for reading and writing files.
It provides extensive support for generating pseudorandom numbers and includes
an assortment of probability distributions.

@@

% https://numpy.org/neps/nep-0016-abstract-array.html
% https://numpy.org/neps/nep-0018-array-function-protocol.html
% https://numpy.org/neps/nep-0022-ndarray-duck-typing-overview.html
% https://numpy.org/neps/nep-0030-duck-array-protocol.html
% https://numpy.org/neps/nep-0037-array-module.html

>>>>

\begin{figure*}
  \centering
  \includegraphics[width=\textwidth]{static/sketches/array-concepts}
  \caption{\fixme{PLACEHOLDER: This figure needs some text to describe
  it}
   }
  \label{fig:array-concepts}
\end{figure*}

@@

The NumPy array consists of a pointer to memory, along with meta-data used to
interpret the data stored there.
Computer memory stores elements linearly and has no notion
of higher-dimensions, but we can interpret it as an $n$-dimensional array by also
providing the array's shape, strides (the number of bytes to advance in memory to skip
along rows, columns, etc.), and data-type (see Figure~\ref{fig:array-concepts}).

Data types (dtypes for short) tell us more about the elements stored in a NumPy
array.
Arrays can hold, e.g., numerical data (of lower and higher precision), strings,
timestamps, or Python objects, each identified by its own dtype.

Once data is viewed as an $n$-dimensional array, more abstract operations such
as reductions (e.g., summation across columns, see Figure~\ref{fig:reductions}) products, and so forth, are
defined.

Users predominantly interact with NumPy arrays using {\em indexing} (to access sub-arrays or individual elements), {\em operators} (e.g., $+$, $-$, $\times$, and $@$, for matrix multiplication), together with {\em utility functions}; together, these provide an easily readable, expressive, high-level API (Application Programming Interface) for array manipulation.

NumPy includes numerous functions that perform element-wise and reduction operations on arrays, including arithmetic, statistics, and trigonometry.  They aim to loop over array elements near-optimally, taking into consideration, e.g., strides, in order to best utilize the computer's fast cache memory.  Vectorized operations that would take many tens of lines to express in languages such as C can often be implemented as a single, clear Python expression.

Other utility functions include creating, reshaping, concatenating, and padding arrays; searching, sorting and counting data; and reading and writing files.  NumPy provides extensive support for generating pseudorandom numbers and includes an assortment of probability distributions. It also performs accelerated linear algebra, utilising one of several backends such as OpenBLAS and Intel MKL optimized for the CPUs at hand.

%Suppose the matrix \code{A} is a 2-d array and the vector \code{x} is a 1-d array.  Then \code{A @ x} computes %the matrix-vector product $Ax$ and \code{x @ A} computes the vector-matrix product $x^\top A$.  This matrix %multiplication operation extends to higher-dimensions by treating the array as a stack of matrices.

Finally, a powerful array feature that affects the way NumPy indexes and combines arrays, is {\em broadcasting}. When performing an operation (such as addition) on two arrays, one intuitively expect that these arrays should be required to have the same shape.  However, through broadcasting, NumPy allows the dimensions to differ, while still producing results that appeal to intuition.  A trivial example is the addition of a scalar value to an array, but it also generalizes to more complex examples such as scaling each column of an array, or generating a grid of coordinates.  In broadcasting, one or both arrays are virtually duplicated (that is, without copying any data in memory), so that the shapes of the operands match (see Figure~\ref{fig:array-concepts}).  Broadcasting is also applied whenever an array is indexed using arrays of indices (see Figure~\ref{fig:broadcasting}).

@@

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{static/sketches/reductions}
  \caption{\fixme{Reducing an array along one of several axes. Note:
  may want to fix to have right-hand axis system.}
   }
  \label{fig:reductions}
\end{figure}

@@

% https://numpy.org/neps/nep-0016-abstract-array.html
% https://numpy.org/neps/nep-0018-array-function-protocol.html
% https://numpy.org/neps/nep-0022-ndarray-duck-typing-overview.html
% https://numpy.org/neps/nep-0030-duck-array-protocol.html
% https://github.com/numpy/numpy/blob/a111b551ae940d7d5f8523fef1cf3589c6ba00a0/doc/neps/nep-0033-extensible-dtypes.rst
% https://numpy.org/neps/nep-0037-array-module.html
